---
title: "Probability Distributions"
format: html
---

## Part 1: Anatomy of a probability distribution

For each of the following random variables, describe:

- What are the events that could possibly happen (sample space)?
- How probable is it that a specific event will happen (probability distribution)?

### Example 1: Coin flip

sample space: 
probability distribution: 



### Example 2: Throw a six-sided die

sample space: 
probability distribution:



### Example 3: The sum of throwing 2 dice

sample space: 
probability distribution:



### Example 4: The height of a student picked at random

sample space: 
probability distribution:



### Example 5: The number of emails you will get today

sample space: 
probability distribution:

---

## Part 2: Match to distributions

Match each of the following scenarios to the most appropriate probability distribution (Uniform, Normal, Bernoulli, Beta, Binomial, Poisson, Exponential, Student's t, Chi-Squared, F-distribution):

1. The rate/percentage of success 
    distribution: 

2. The IQ scores
    distribution: 

3. The number of emails received in one day
    distribution: 

4. The probability of winning the lottery among lottery players
    distribution: 

5. The ratio of variances
    distribution: 

6. The distribution of squared errors
    distribution: 

7. The number of people cured among the treatment group of 10 people
    distribution: 

8. The duration between two students entering MSA
    distribution: 

9. A coin landing on heads
    distribution: 

10. The distribution of an estimated mean (unknown population std.)
    distribution: 



---

## Part 3: The 4 types of questions to ask a probability distribution

R provides four functions for every distribution: 

- `d___()` (density), 
- `p___()` (cumulative probability), 
- `q___()` (quantile), 
- `r___()` (random samples).

for example (dnorm(), ppois(), qbeta(), rbinom())



Identify which type of question (d, p, q, or r) each of the following corresponds to:

1. Pick a random card out of a 52-card deck: 
    it's a "x____ question"

2. What's the probability of getting a 6 when throwing a die?
    it's a "x____ question"

3. What IQ would you need to have for 30% of the population to have a lower IQ than you?
    it's a "x____ question"

4. How do you calculate a p-value?
    it's a "x____ question"

5. What percentage of people have an IQ larger than 130?
    it's a "x____ question"

6. What's the probability that a psychology student earns between 5,000€ and 6,000€?
    it's a "x____ question"

7. Guess a number between 1 and 10
    it's a "x____ question"




---

## Part 4: Simulate IQ scores

- Generate 10 random numbers from a normal distribution with mean = 100 and sd = 15.
- Calculate the mean of those IQ numbers. What would you expect?
- Do the same with 100, 1000, and 10,000 numbers. What changes?

```{r}
# Your code here

```



---

## Part 5: Test the Central Limit Theorem

Write code to:

- Simulate the throw of *n* dice
- Compute the sum of those *n* dice
- Store that value and repeat 1,000 times
- Plot the distribution of those sums
- Does it look "normal"?

Try with n = 1, 2, and 50 dice. What do you observe?

```{r}
# Your code here

```
