---
title: "Standardized Regression Coefficients"
format: html
---



## About

This notebook covers **standardized regression coefficients** — what they are, how to compute them, and why they matter.

We will focus on three key topics:

1. **Standardising variables** — z-scoring predictors and outcomes before fitting a linear model.
2. **Standardized vs unstandardized coefficients** — how the two relate, when they differ, and which one to report.
3. **Simple regression special case** — showing that the standardized coefficient equals the Pearson correlation $r$ when there is only one predictor.

```{r}
library(tidyverse)
library(GGally)
```


---

## Data

We reuse the simulated salary/age dataset from the previous labs. The true model is:

$$\text{salary} = 2500 + (\text{age} - 18) \times 120 + \varepsilon$$

We also include **years of experience**, which is correlated with age.

```{r}
set.seed(42)
n <- 100

age        <- runif(n, min = 18, max = 65)
err        <- rnorm(n, 0, 1000)
salary     <- 2500 + (age - 18) * 120 + err
experience <- pmax((age - 18) * 0.8 + rnorm(n, 0, 3), 0)

sim_data <- data.frame(age, experience, salary)
head(sim_data)
```


---

## Part 1 — What are standardized coefficients?

### The problem with raw (unstandardized) coefficients

In a regression model, each coefficient is expressed in the **original units** of that variable. For instance, in our model:

$$\widehat{\text{salary}} = \beta_0 + \beta_1 \cdot \text{age} + \beta_2 \cdot \text{experience}$$

- $\beta_1$ is in dollars-per-year-of-age
- $\beta_2$ is in dollars-per-year-of-experience

These are on different scales. If we were comparing age (measured in years, range ≈ 18–65) against some variable measured in, say, months or dollars, a direct comparison of $\beta_1$ vs $\beta_2$ would be meaningless.

**Standardized coefficients** solve this by putting everything in units of standard deviations: a standardized coefficient $\beta^*$ tells you *how many standard deviations $y$ changes for a one-SD increase in $x$*. This makes coefficients comparable across predictors and across studies.


### Step 1 — Standardise the variables (z-score)

Z-scoring a variable $x$ means: subtract its mean and divide by its standard deviation.

$$z_x = \frac{x - \bar{x}}{s_x}$$

After z-scoring, every variable has mean 0 and SD 1. 


You can z-score a variable as follows

```{r}

age_z <- ( sim_data$age - mean(sim_data$age) ) / sd(sim_data$age)

cat("Mean of standardized `age`: ",  round(mean(age_z), 4) )
cat("\n")
cat("SD of standardized `age`: ",  round(sd(age_z), 4) )

```


Instead of doing this manually and separately for each variable, we can use `scale()` for z-scoring and use the tidyverse tools to apply this to all variables at once:

```{r}
sim_data_z <- sim_data |>
  mutate(across(everything(), ~ as.numeric(scale(.))))

# Verify: means should be ~0, SDs should be ~1
sim_data_z |>
  summarise(across(everything(), list(mean = mean, sd = sd))) |>
  pivot_longer(everything(), names_to = c("variable", "stat"), names_sep = "_(?=[^_]+$)") |>
  pivot_wider(names_from = stat, values_from = value)
```


---

## Part 2 — Simple regression: standardized coefficient = correlation

Let's start with the simplest case: **one predictor**.

We can fit the same model on either the raw data or on the standardized data:

```{r}
# --- Unstandardized simple regression ---
m_raw <- lm(salary ~ age, data = sim_data)

# --- Standardized simple regression (z-scored data) ---
m_std <- lm(salary ~ age, data = sim_data_z)

```



```{r}
summary(m_raw)
```



```{r}
summary(m_std)
```


When comparing the results of these two model fits, you can see that 

  a) the estimated coefficients are different
  b) the overall model statistics (i.e., R-squared, F, p-value) are identical



Does this make sense to you?





By transforming the variables we can facilitate the interpretation of the coefficient or give them properties we believe are useful.


In a simple linear regression of $y$ on $x$, the standardized slope equals the Pearson correlation between $x$ and $y$:

$$\beta^*_{yx} = r_{xy}$$

This is a key insight. Let's verify it with our data.


```{r}
# Pearson correlation between age and salary
r_age_salary <- cor(sim_data$age, sim_data$salary)

# Standardized slope from the z-scored model
beta_std <- coef(m_std)["age"]

cat("Pearson r(age, salary)         =", round(r_age_salary, 6), "\n")
cat("Standardized beta (age->salary) =", round(beta_std,     6), "\n")
cat("Are they equal?", isTRUE(all.equal(r_age_salary, beta_std, tolerance = 1e-10)), "\n")
```


### Why does this equivalence hold?

The formula for the OLS slope in a simple regression is:

$$\hat{\beta}_1 = \frac{\text{Cov}(x, y)}{\text{Var}(x)}$$

When both $x$ and $y$ are standardized (SD = 1), $\text{Var}(x) = 1$ and $\text{Cov}(x, y) = r_{xy}$, so:

$$\hat{\beta}^*_1 = \frac{r_{xy}}{1} = r_{xy}$$

This equivalence breaks down as soon as you add a second predictor (multiple regression), because the slope for each predictor is then adjusted for the others. In that case $\beta^* \neq r_{xy}$ in general — which is precisely what makes multiple regression more informative than a set of pairwise correlations.

**Q: If the standardized coefficient equals $r$ in simple regression, what does the intercept equal after standardising? Check this from the output above.**

> Write your answer here:

____


---

## Part 3 — Converting between standardized and unstandardized coefficients

You don't always need to re-run a model on z-scored data. Given the unstandardized slope $\hat{\beta}_1$ and the standard deviations of $x$ and $y$, you can convert:

$$\beta^*_1 = \hat{\beta}_1 \cdot \frac{s_x}{s_y}$$

And the reverse:

$$\hat{\beta}_1 = \beta^*_1 \cdot \frac{s_y}{s_x}$$

Let's verify this manually:

```{r}
beta_raw <- coef(m_raw)["age"]   # unstandardized slope
s_x      <- sd(sim_data$age)
s_y      <- sd(sim_data$salary)

# Convert raw -> standardized
beta_std_manual <- beta_raw * (s_x / s_y)

cat("Standardized beta (z-scored model)    =", round(coef(m_std)["age"],  6), "\n")
cat("Standardized beta (manual conversion) =", round(beta_std_manual,      6), "\n")
```


---

## Part 4 — Multiple regression: standardized vs unstandardized

### Fitting and comparing the models

Now let's fit the full multiple regression with both `age` and `experience`, in both raw and standardized form.

```{r}
# Unstandardized multiple regression
m_multi_raw <- lm(salary ~ age + experience, data = sim_data)

# Standardized multiple regression
m_multi_std <- lm(salary ~ age + experience, data = sim_data_z)

cat("=== Unstandardized ===\n")
summary(m_multi_raw)$coefficients

cat("\n=== Standardized ===\n")
summary(m_multi_std)$coefficients
```

### Reading the standardized output

In the standardized model:

- The **intercept is 0** (both $x$ and $y$ are centred at 0 after z-scoring).
- Each slope tells you: *a 1-SD increase in this predictor is associated with a $\beta^*$-SD change in salary, holding the other predictor constant.*
- You can now **directly compare** the coefficients: whichever has the larger absolute value has the stronger unique association with salary.
- BUT: these coefficients depend on the variance of the variables (e.g., the range of ages you included in your data)



**Q: Which predictor has a larger standardized coefficient — `age` or `experience`? Is this surprising given that salary was simulated to depend only on `age`? Why might `experience` still have a nonzero coefficient?**

> Write your answer here:

____


### Standardized coefficients do NOT equal correlations in multiple regression

In simple regression we showed $\beta^* = r$. In multiple regression this no longer holds — the coefficients are adjusted for the other predictors. Let's confirm:

```{r}
r_age_sal  <- cor(sim_data$age,        sim_data$salary)
r_exp_sal  <- cor(sim_data$experience, sim_data$salary)
beta_age   <- coef(m_multi_std)["age"]
beta_exp   <- coef(m_multi_std)["experience"]

comparison <- tibble(
  predictor        = c("age", "experience"),
  `zero-order r`   = c(r_age_sal, r_exp_sal),
  `standardized b` = c(beta_age,  beta_exp),
  difference       = c(r_age_sal - beta_age, r_exp_sal - beta_exp)
)

print(comparison)
```

The difference between $r$ and $\beta^*$ reflects the adjustment for the shared variance between `age` and `experience`. Because the two predictors are correlated, including both changes each coefficient relative to the zero-order correlation.


---

## Part 5 — Visual comparison

### Coefficient plot: raw vs standardized

A useful way to communicate regression results is a **coefficient plot** — one dot per predictor, with its confidence interval. In the standardized version all predictors are on the same scale, making comparisons fair.

```{r}
coef_plot_data <- function(model, label) {
  ci    <- confint(model)
  coefs <- coef(model)
  tibble(
    model     = label,
    predictor = names(coefs),
    estimate  = unname(coefs),
    lower     = unname(ci[, 1]),
    upper     = unname(ci[, 2])
  ) |> filter(predictor != "(Intercept)")
}

plot_dat <- bind_rows(
  coef_plot_data(m_multi_raw, "Unstandardized"),
  coef_plot_data(m_multi_std, "Standardized")
)

ggplot(plot_dat, aes(x = estimate, y = predictor, xmin = lower, xmax = upper)) +
  geom_pointrange(size = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ model, scales = "free_x") +
  labs(
    title = "Regression coefficients: unstandardized vs standardized",
    x     = "Coefficient estimate (± 95% CI)",
    y     = NULL
  ) +
  theme_minimal()
```

**Q: In the unstandardized panel the coefficients for `age` and `experience` look very different in magnitude. In the standardized panel they look much closer. What does this tell you about the importance of scale when comparing predictors?**

> Write your answer here:

____


---

## Part 6 — Model fit is identical regardless of standardisation

Standardising the variables does **not** change the model fit: R², adjusted R², and the F-statistic are all the same. Let's confirm:

```{r}
fit_comparison <- tibble(
  model       = c("Unstandardized", "Standardized"),
  R2          = c(summary(m_multi_raw)$r.squared,
                  summary(m_multi_std)$r.squared),
  adj_R2      = c(summary(m_multi_raw)$adj.r.squared,
                  summary(m_multi_std)$adj.r.squared),
  F_statistic = c(summary(m_multi_raw)$fstatistic["value"],
                  summary(m_multi_std)$fstatistic["value"])
)

print(fit_comparison)
```

R² and the F-statistic are identical — standardisation is a linear rescaling, so it leaves the geometry of the regression unchanged. Only the *units* of the coefficients change, not what the model explains.


---

## Summary

| Property | Unstandardized $\hat{\beta}$ | Standardized $\beta^*$ |
|---|---|---|
| Units | Original units of $y$ per unit of $x$ | SDs of $y$ per SD of $x$ |
| Intercept | Mean of $y$ when all $x = 0$ | Always 0 |
| Comparable across predictors? | No | Yes |
| Simple regression: equals $r_{xy}$? | No | **Yes** |
| Multiple regression: equals $r_{xy}$? | No | No (adjusted for other predictors) |
| Changes model fit (R², F)? | — | No |

The key takeaway: **standardized coefficients tell you about relative importance within a model; they do not change what the model predicts or how well it fits**.



## Exercise

The `attitude` dataset (built into R) contains survey responses from 30 employees on seven variables:

| Variable | Description |
|---|---|
| `rating` | Overall rating of the company's work climate |
| `complaints` | Handling of employee complaints |
| `privileges` | Employee privileges |
| `learning` | Opportunity for learning |
| `raises` | Opportunity for raises |
| `advance` | Opportunity for advancement |
| `critical` | How critical the supervisor is |


```{r}
ggpairs(attitude, progress = FALSE)
```



Compute a linear regression of rating as a function of raises using a) the raw data and b) the standardized (z-scored) data.

```{r}
# linear model on the raw data
# write your code here:

```

```{r}
# linear model on the z-scored data
# write your code here:

```


Compare and interpret the coefficients of the two models. Which one is the best?

> Write your answer here:


Which coefficient is equivalent to the linear correlation between `rating` and `raises`? 

> Write your answer here:


Write code to test that this is indeed correct.


```{r}
# write your coce here:

```

